---
title: "🏆 The GTA IV Green Transit Awards 2025 🏆"
author: "Shubh Goyal"
date: today
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    code-summary: "Show Code"
    code-tools: true
    doctype: html
engine: knitr 
execute:
  echo: true  # Show code
  eval: true  # Ensure code runs
  warning: false
  message: false
editor: visual
markdown: 
  wrap: 68
---

# 🚍 The 2025 Clean Transit Awards

**Honoring the Green, the Lean, and the Gas-Guzzling Machines! 🌱🔥**

New York, NY – Public transit plays a crucial role in reducing carbon footprints, but not all agencies are created equal. Some transit systems have become champions of eco-efficiency, while others… well, let’s just say they have room for improvement.

After analyzing emissions data from transit agencies nationwide, we proudly present the 2025 Clean Transit Awards, recognizing the best (and worst) in sustainable transit. These awards highlight Small, Medium, and Large agencies that have made a measurable impact on their carbon footprint—or left tire marks on the environment.

So, without further ado, let’s reveal the greenest, the most improved, and, of course, the biggest gas guzzlers!

## 🏆 Greenest Transit Agency – The MVP (Most Valuable Pollution-Reducer) Award 🌍💨

This prestigious award goes to the transit agencies that produce the least emissions per passenger trip and per mile traveled. In other words, these winners are basically the Tesla Model S of buses—except way more affordable and accessible.

**By Passenger Trips:**

🥇**Large Agency Winner:** New Jersey Transit Corporation

🥇**Medium Agency Winner:** King County, dba: King County Metro

🥇**Small Agency Winner**: Small Altoona Metro Transit, dba: AMTRAN

**By Passenger Mile:**

🥇**Large Agency Winner**:🚆New Jersey Transit Corporation,

🥇**Medium Agency Winner:** King County, dba: King County Metro**,**

🥇**Small Agency Winner:** Small Altoona Metro Transit, dba: AMTRA**,**

## 🌍 Most Emissions Avoided – The "Carbon Buster" Award 🚫💨

"If pollution had an arch-nemesis, it would be these guys."

Imagine a world where all transit riders suddenly switched to single-passenger cars. 🚗🚗🚗 Yikes! The traffic jams would be longer than your uncle’s conspiracy theories at Thanksgiving, and emissions would skyrocket.

This award goes to the transit agencies that saved us from the most pollution by preventing countless solo car trips.

🥇 **Large Agency Winner**: MTA New York City Transit 3441010874 avoided

🥇 **Medium Agency Winner**: Metro-North Commuter Railroad Company, dba: MTA Metro-North Railroad 413801729 avoided

🥇 **Small Agency Winner**: Hampton Jitney, Inc. 1333301 avoided

## Most Improved Green Agency – The "Comeback Kid" Award 🔄💚

"From gas-guzzler to eco-hustler!"

This award goes to the agency that has made the biggest leap toward sustainability—not by comparing to past-year data, but by seeing who improved the most compared to their peers.

🥇 **Large Agency Winner**: New Jersey Transit Corporation Reduced from 0.1260740 to 0.0002060647 per mile

🥇 **Medium Agency Winner**: King County, dba: King County Metro Reduced from 0.2200267 to 0.0001889437 per mile

🥇 **Small Agency Winner**: Altoona Metro Transit, dba: AMTRAN – Reduced from 0.3648839 to 0.0015635784 per mile


## ⛽ Biggest Gas Guzzler – The "Oh No, Fix That!" Award💀

"Not all heroes wear capes... and not all transit agencies are heroes."Not every agency can be a green machine. This award "honors" those transit agencies that burn through fuel like it’s going out of style, producing the highest emissions per passenger mile. (Spoiler alert: This is NOT the award you want to win.)
Chicago Transit Authority	162.6084		
Medium	West Virginia University, dba: Personal Rapid Transit	645.1318		
Small	Pennsylvania Department of Transportation	273.8586	 

🥇 **Large Agency Winner**: SChicago Transit Authority	162.6084 kg CO₂ per mile

🥇 **Medium Agency Winner**: West Virginia University, dba: Personal Rapid Transit	645.1318 kg CO₂ per mile

🥇 **Small Agency Winner**: Pennsylvania Department of Transportation	273.8586 CO₂ per mile


While some agencies are already champions of green transit, others have some serious catching up to do. The 2025 Clean Transit Awards highlight the best in eco-friendly transportation while also shining a spotlight on areas that need improvement. 🚆 The takeaway? Cities that invest in sustainable transit are making a real impact on emissions. Meanwhile, gas guzzlers (we’re looking at you, WVU!) still have some work to do. For a full breakdown of the data, check out the Appendix in our official report. Spoiler alert: numbers don’t lie, and neither do emissions reductions!

# APPENDIX - METHODOLOGY AND ANALYSIS

## 📊 Data Import

We started by acquiring data from EIA State Electricity Profiles, which we will use to estimate the environmental impact of the electricity used to run certain transit systems. 🚗

#### State Electricity Profiles 🏛️

For purposes of this analysis, we will assume that all electricity generation in a state is fungible. That is, we will assign the average generation emissions to all-electric transit in a state, even if the transit authority officially has a “green supply” agreement in place.
```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(knitr)
library(DT)
library(tidyr)
library(readr)
library(stringr) # Needed for str_replace_all()
library(purrr) # Needed for map2()
library(quarto) # Ensures Quarto execution works
```

```{r}
# Load libraries
ensure_package <- function(pkg){
    pkg <- as.character(substitute(pkg))
    options(repos = c(CRAN = "https://cloud.r-project.org"))
    if(!require(pkg, character.only=TRUE)) install.packages(pkg)
    stopifnot(require(pkg, character.only=TRUE))
}
ensure_package(stringr)  # Ensure it's installed
ensure_package(dplyr)
ensure_package(httr2)
ensure_package(rvest)
ensure_package(datasets)
ensure_package(purrr)
ensure_package(DT)

get_eia_sep <- function(state, abbr){
    state_formatted <- str_to_lower(state) |> str_replace_all("\\s", "")
    
    dir_name <- file.path("data", "mp02")
    file_name <- file.path(dir_name, state_formatted)
    
    dir.create(dir_name, showWarnings=FALSE, recursive=TRUE)
    
    if(!file.exists(file_name)){
        BASE_URL <- "https://www.eia.gov"
        REQUEST <- request(BASE_URL) |> 
            req_url_path("electricity", "state", state_formatted)
    
        RESPONSE <- req_perform(REQUEST)
    
        resp_check_status(RESPONSE)
        
        writeLines(resp_body_string(RESPONSE), file_name)
    }
    
    TABLE <- read_html(file_name) |> 
        html_element("table") |> 
        html_table() |>
        mutate(Item = str_to_lower(Item))
    
    if("U.S. rank" %in% colnames(TABLE)){
        TABLE <- TABLE |> rename(Rank = `U.S. rank`)
    }
    
    CO2_MWh <- TABLE |> 
        filter(Item == "carbon dioxide (lbs/mwh)") |>
        pull(Value) |> 
        str_replace_all(",", "") |>
        as.numeric()
    
    PRIMARY <- TABLE |> 
        filter(Item == "primary energy source") |> 
        pull(Rank)
    
    RATE <- TABLE |>
        filter(Item == "average retail price (cents/kwh)") |>
        pull(Value) |>
        as.numeric()
    
    GENERATION_MWh <- TABLE |>
        filter(Item == "net generation (megawatthours)") |>
        pull(Value) |>
        str_replace_all(",", "") |>
        as.numeric()
    
    data.frame(CO2_MWh               = CO2_MWh, 
               primary_source        = PRIMARY,
               electricity_price_MWh = RATE * 10, # / 100 cents to dollars &
               # * 1000 kWh to MWH 
               generation_MWh        = GENERATION_MWh, 
               state                 = state, 
               abbreviation          = abbr
    )
}

EIA_SEP_REPORT <- map2(state.name, state.abb, get_eia_sep) |> list_rbind()

ensure_package(scales)
ensure_package(DT)

EIA_SEP_REPORT |> 
    select(-abbreviation) |>
    arrange(desc(CO2_MWh)) |>
    mutate(CO2_MWh = scales::number(CO2_MWh, big.mark=","), 
           electricity_price_MWh = scales::dollar(electricity_price_MWh), 
           generation_MWh = scales::number(generation_MWh, big.mark=",")) |>
    rename(`Pounds of CO2 Emitted per MWh of Electricity Produced`=CO2_MWh, 
           `Primary Source of Electricity Generation`=primary_source, 
           `Average Retail Price for 1000 kWh`=electricity_price_MWh, 
           `Total Generation Capacity (MWh)`= generation_MWh, 
           State=state) |>
    kable(caption = "EIA State Electricity Profile Report")
```
🌍 Here, we have collected the effective emissions per MWh2⚡ as well as price and total state-wide generation capacity. We will use this data to compare the emissions of different transit modalities. We next turn to the National Transit Database to get information on various public transit authorities. Before we do so, however, let’s explore the SEP data a bit first.🔬📉

## 🔍 Initial Analysis of SEP Data ⚖️

Let’s answer some burning (fossil-free, we hope!) questions: 🚇

### 1. Which state has the most expensive retail electricity? 💰⚡

Which state will make you cry when you get your electricity bill? 😭 Let's find out!


```{r}
EIA_SEP_REPORT |> 
  arrange(desc(electricity_price_MWh)) |> 
  select(state, electricity_price_MWh) |> 
  head(1) |> 
  kable(caption = "Most Expensive Retail Electricity")
```

### 2. Which state has the ‘dirtiest’ electricity mix? ☠️🌫️

Which state’s power plants are puffing out CO₂ like an angry dragon? 🐉🔥 Stay tuned for some shocking revelations!


```{r}
EIA_SEP_REPORT |> 
  arrange(desc(CO2_MWh)) |> 
  select(state, CO2_MWh) |> 
  head(1) |> 
  kable(caption = "Dirtiest electricity mix")
```

### 3. On average, how many pounds of CO2 are emitted per MWh of electricity produced in the US?🏭

Time for some serious math! 📊 We’ll calculate the national average CO₂ emissions per MWh using a **weighted average** because we like our stats accurate.🎯


```{r}
# Calculate weighted average CO2 emissions
weighted_avg_CO2 <- sum(EIA_SEP_REPORT$CO2_MWh * EIA_SEP_REPORT$generation_MWh) / 
  sum(EIA_SEP_REPORT$generation_MWh)

data.frame(Weighted_Avg_CO2 = weighted_avg_CO2) |> 
  kable(col.names = c("Weighted Average CO2 Emissions (lbs/MWh)"),
        caption = "Weighted Average CO2 Emissions Across All States")

```

### 4. 🦄 What is the rarest primary energy source in the US? What is the associated cost of electricity and where is it used?🧐

What’s the mythical unicorn of US energy sources? 🦄✨ And where is it hiding? Let’s track it down and see how much it costs! 💸


```{r}
EIA_SEP_REPORT |> 
  count(primary_source) |> 
  arrange(n) |> 
  slice(1) |> 
  inner_join(EIA_SEP_REPORT, by = "primary_source") |> 
  select(state, primary_source, electricity_price_MWh) |> 
  kable(caption = "Primary Energy Source with Lowest Occurrence")
```

### 5. My home state, Texas, has a reputation as being the home of “dirty fossil fuels” while NY has a reputation as a leader in clean energy. How many times cleaner is NY’s energy mix than that of Texas?🌱🔥

Is Texas really the king of fossil fuels? 👑 And does New York truly deserve its green halo? 😇 Let’s compare their energy mixes and see **how many times cleaner NY is compared to TX**! 📊⚖️

```{r}
NY_CO2 <- EIA_SEP_REPORT |> filter(state == "New York") |> pull(CO2_MWh)
TX_CO2 <- EIA_SEP_REPORT |> filter(state == "Texas") |> pull(CO2_MWh)

cleaner_ratio <- TX_CO2 / NY_CO2

tibble(Metric = "Cleaner Ratio (TX to NY)", Value = cleaner_ratio) |> 
  kable(caption = "Texas vs. New York CO2 Emissions Ratio")
```

Fun fact: New York’s energy mix is cleaner than Texas’s by a factor of **1.637931**. That’s right, the Lone Star State might want to take notes from the Big Apple! 🍏⚡

## Importing 2023 Annual Database Energy Consumption Report ⚡📊

To analyze the energy consumption data for public transit agencies, we first need to import the **2023 Annual Database Energy Consumption** report. This Dataset is provided in an **Excel spreadsheet (.xlsx)** format. We will use the readxl package in R to load it.

```{r}
ensure_package(readxl)
# Create 'data/mp02' directory if not already present
DATA_DIR <- file.path("data", "mp02")
dir.create(DATA_DIR, showWarnings=FALSE, recursive=TRUE)

NTD_ENERGY_FILE <- file.path(DATA_DIR, "2023_ntd_energy.xlsx")

if(!file.exists(NTD_ENERGY_FILE)){
    DS <- download.file("https://www.transit.dot.gov/sites/fta.dot.gov/files/2024-10/2023%20Energy%20Consumption.xlsx", 
                  destfile=NTD_ENERGY_FILE, 
                  method="curl")
    
    if(DS | (file.info(NTD_ENERGY_FILE)$size == 0)){
        cat("I was unable to download the NTD Energy File. Please try again.\n")
        stop("Download failed")
    }
}

NTD_ENERGY_RAW <- read_xlsx(NTD_ENERGY_FILE)
```
```{r}
### Data Cleaning 🧹


ensure_package(tidyr)
to_numeric_fill_0 <- function(x){
    x <- if_else(x == "-", NA, x)
    replace_na(as.numeric(x), 0)
}

NTD_ENERGY <- NTD_ENERGY_RAW |> 
    select(-c(`Reporter Type`, 
              `Reporting Module`, 
              `Other Fuel`, 
              `Other Fuel Description`)) |>
    mutate(across(-c(`Agency Name`, 
                     `Mode`,
                     `TOS`), 
                  to_numeric_fill_0)) |>
    group_by(`NTD ID`, `Mode`, `Agency Name`) |>
    summarize(across(where(is.numeric), sum), 
              .groups = "keep") |>
    mutate(ENERGY = sum(c_across(c(where(is.numeric))))) |>
    filter(ENERGY > 0) |>
    select(-ENERGY) |>
    ungroup()

```

### **🎲 The Wheel of Randomness! 🎰**

**Ladies and gentlemen, it's time for the Random Transit Energy Lottery! 🎉🎟️ We’ll now select 10 lucky transit energy records at random! Who will make the cut? Who will be left behind? Let's spin the wheel! 🔄✨**

```{r}

# Sample 10 random rows and then pass them to kable()
NTD_ENERGY |> 
  slice_sample(n = 10) |> 
  kable(caption = "Random Sample of NTD_ENERGY Data")

```

## 🚇 Recoding Transit Modes 🚋🎭

🚧 **Warning: Mysterious Codes Ahead!** 🚧

Before we can make sense of the **Mode** column, we need to crack the code! 🕵️‍♂️ First, let's uncover the unique **Mode codes** in our data using `distinct()`. Then, we consult the sacred texts (a.k.a. the NTD website 📜) to decipher their meanings. 🧐

### 🏗️ Time for a Makeover! 🖌️

The **Mode** column needs a glow-up! ✨ We’re about to give those cryptic abbreviations some **real names** so they’re easier to work with. Ready? Let’s do this! 🎬

```{r}
NTD_ENERGY <- NTD_ENERGY |>
  mutate(Mode = case_when(
    Mode == "HR"  ~ "Heavy Rail", 
    Mode == "LR"  ~ "Light Rail", 
    Mode == "CR"  ~ "Commuter Rail", 
    Mode == "MB"  ~ "Motor Bus", 
    Mode == "TB"  ~ "Trolleybus", 
    Mode == "TR"  ~ "Heritage Streetcar", 
    Mode == "SR"  ~ "Modern Streetcar", 
    Mode == "VP"  ~ "Vanpool", 
    Mode == "DR"  ~ "Demand Response", 
    Mode == "FB"  ~ "Ferryboat", 
    Mode == "MG"  ~ "Monorail/Automated Guideway", 
    Mode == "PB"  ~ "Publico", 
    Mode == "CB"  ~ "Commuter Bus",
    Mode == "RB"  ~ "Rapid Bus",
    Mode == "AR"  ~ "Aerial Tramway",
    Mode == "YR"  ~ "Hybrid Rail",
    Mode == "IP"  ~ "Inclined Plane Railway",
    Mode == "CC"  ~ "Cable Car",
    TRUE          ~ "Unknown"
  ))

NTD_ENERGY |> distinct(Mode) |>
kable(caption = "Distinct Modes of Transport")

```

## 2023 Annual Database Service by Agency 📊🚇

🚨 **Breaking News!** 🚨 The latest **2023 Service by Agency** report has just dropped! 🎤📉 We’re diving in to extract the juicy details about passenger trips across different transit services. 🚆🚌

```{r}
NTD_SERVICE_FILE <- file.path(DATA_DIR, "2023_service.csv")
if(!file.exists(NTD_SERVICE_FILE)){
    DS <- download.file("https://data.transportation.gov/resource/6y83-7vuw.csv", 
                  destfile=NTD_SERVICE_FILE, 
                  method="curl")
    
    if(DS | (file.info(NTD_SERVICE_FILE)$size == 0)){
        cat("I was unable to download the NTD Service File. Please try again.\n")
        stop("Download failed")
    }
}

NTD_SERVICE_RAW <- read_csv(NTD_SERVICE_FILE)
```

📢 **Will our dataset arrive safely, or will it ghost us? 👻** Only time (and a stable internet connection) will tell! ⏳

## Clean-Up Time! 🧹✨

```{r}
NTD_SERVICE <- NTD_SERVICE_RAW |>
    mutate(`NTD ID` = as.numeric(`_5_digit_ntd_id`)) |> 
    rename(Agency = agency, 
           City   = max_city, 
           State  = max_state,
           UPT    = sum_unlinked_passenger_trips_upt, 
           MILES  = sum_passenger_miles) |>
    select(matches("^[A-Z]", ignore.case=FALSE)) |>
    filter(MILES > 0)
```

🛠️ **Boom! Now our dataset is sparkling clean!** ✨ It’s time to put these numbers to work! 💪

## 🛥️ Exploring NTD Service Data - Let’s Dive In!

Alright, folks! Time to put on our data analyst hats and explore some seriously juicy transit numbers from the **NTD_SERVICE** Dataset. We're going to answer some fun questions that might just reveal who’s cruising on the green lane of transit efficiency and who's still stuck in the slow lane. Let’s see how this data speeds up the fun! 😎

### 1. Which transit service has the most UPT annually? 🏆

Which transit service is the busiest, with the most Unlinked Passenger Trips (UPT) annually? Spoiler alert: It’s probably not the one with the tiniest buses.🚌

```{r}
NTD_SERVICE |>
  filter(!is.na(UPT)) |>
  arrange(desc(UPT)) |>
  select(Agency, UPT) |>
  head(1) |>
  kable(caption = "Most UTP Annually")
```

### 2. What is the average trip length of a trip on MTA NYC? 🗽

Time for some serious subway sleuthing! 🕵️‍♂️ How long is the average trip on the MTA NYC? Are New Yorkers making quick, subway sprints, or are they getting comfy for a longer ride? We’re on a mission to figure out just how far these trips stretch, so hop on board! 🚇⏳

```{r}
NTD_SERVICE |>
  filter(grepl("MTA", Agency, ignore.case = TRUE)) |>
  summarise(Average_Trip_Length = sum(MILES, na.rm = TRUE) / sum(UPT, na.rm = TRUE)) |>
  kable(caption = "Average Trip Length On MTA Nyc")
```

### 3. Which transit service in NYC has the longest average trip length? 🚋

Which MTA service in New York City (including all five boroughs, yes, even Brooklyn 🗽) has the longest average trip length? Let’s just say, it’s probably not the one where the subway rats are the most frequent passengers. 🐀

P.S. You’ll want to be extra careful with that City column; we wouldn’t want to accidentally send our buses on a detour to Brooklyn when they should be cruising through Manhattan!

```{r}
NYC_SERVICES <- NTD_SERVICE |>
  filter(State == "NY", MILES > 0, UPT > 0) |>
  mutate(Average_Trip_Length = MILES / UPT) |>
  arrange(desc(Average_Trip_Length)) |>
  select(Agency, City, Average_Trip_Length)

head(NYC_SERVICES, 1) |>
kable(caption = "NYC Longest Average Trip Length")
```

### 4. Which state has the fewest total miles travelled by public transit? 🐢

Which state is the turtle in the race, with the fewest total miles travelled by public transit? 🐢 Are some states just cruising along at a leisurely pace, or are they barely moving at all? It’s time to find out who’s not clocking up the miles! 🛑🚗

```{r}
NTD_SERVICE |>
  group_by(State) |>
  summarise(Total_Miles = sum(MILES, na.rm = TRUE)) |>
  arrange(Total_Miles) |>
  head(1) |>
  kable(caption = "Fewest total miles Travelled By Public Transit")
```

### 5. Are all states represented in this data? If no, which ones are missing? The state.name and state.abb objects we used above may be useful here.🚏🗺️

Now for a little mystery – are all states in this data? Or did a few sneak out of the spotlight? 👀 Let’s use the state.name and state.abb to track down any states that may have taken the wrong exit. Looks like it’s time for a little data detective work! 🔍💼

```{r}
all_states <- state.abb  # Vector of all US state abbreviations
present_states <- unique(NTD_SERVICE$State)
missing_states <- setdiff(all_states, present_states)
kable(missing_states, caption = "Missing States")
```

## 🌍 Calculate Emissions – Let’s Crunch Some Carbon! 💨

Time to put on our environmental cap and get down to the nitty-gritty of CO2 emissions! We're about to take a deep dive into some data to calculate the total emissions from different transit agencies and their modes. Spoiler alert: We’re going green – but with a twist! 🌱🔢

### Join the Three Tables – The Ultimate Data Party 🎉

Here’s where we get our hands dirty (but in an eco-friendly way, of course!). We’ll join three tables together like the perfect data trio:

**NTD_SERVICE** 🚍 – It tells us the Agency and Mode (because, let’s face it, transit is all about style, right?)

**NTD_ENERGY** ⚡ – This table’s going to give us the lowdown on all the fuel sources. It’s like fuel’s Wikipedia page, but with more numbers.

**EIA_SEP_REPORT** 📊 – This is where the magic happens! It tells us the CO2 emissions per MWh. Time to add some climate-friendly math into the mix. Once we’ve got these three tables chatting with each other, we’re ready to roll!

```{r}
# Create a mapping of state abbreviations to full names
state_abbrev_to_full <- c(
  "AL" = "Alabama", "AK" = "Alaska", "AZ" = "Arizona", "AR" = "Arkansas", "CA" = "California", "CO" = "Colorado", 
  "CT" = "Connecticut", "DE" = "Delaware", "FL" = "Florida", "GA" = "Georgia", "HI" = "Hawaii", "ID" = "Idaho", 
  "IL" = "Illinois", "IN" = "Indiana", "IA" = "Iowa", "KS" = "Kansas", "KY" = "Kentucky", "LA" = "Louisiana", 
  "ME" = "Maine", "MD" = "Maryland", "MA" = "Massachusetts", "MI" = "Michigan", "MN" = "Minnesota", "MS" = "Mississippi", 
  "MO" = "Missouri", "MT" = "Montana", "NE" = "Nebraska", "NV" = "Nevada", "NH" = "New Hampshire", "NJ" = "New Jersey", 
  "NM" = "New Mexico", "NY" = "New York", "NC" = "North Carolina", "ND" = "North Dakota", "OH" = "Ohio", "OK" = "Oklahoma", 
  "OR" = "Oregon", "PA" = "Pennsylvania", "RI" = "Rhode Island", "SC" = "South Carolina", "SD" = "South Dakota", 
  "TN" = "Tennessee", "TX" = "Texas", "UT" = "Utah", "VT" = "Vermont", "VA" = "Virginia", "WA" = "Washington", 
  "WV" = "West Virginia", "WI" = "Wisconsin", "WY" = "Wyoming"
)

# Fix State Names in NTD_SERVICE before joining
NTD_SERVICE <- NTD_SERVICE |> mutate(State = state_abbrev_to_full[State])

# Now perform the joins
NTD_EMISSIONS <- NTD_SERVICE |>
  select(Agency, State, `NTD ID`) |>
  inner_join(NTD_ENERGY, by = "NTD ID") |>
  inner_join(EIA_SEP_REPORT, by = c("State" = "state"))

# Check if the join worked
dim(NTD_EMISSIONS)  # Should no longer be (0, X)

```

### The Ultimate Emissions Table 🧮

**We’re going to create a super sleek table with all the goods:**

**Agency + Mode Pair –** These are the dynamic duos of transit! 🚍🚆

**State – Every agency has a home**, and we’ll find out which state it’s calling its turf! 🌎

**Fuel Sources** – Because we need to know how those buses are powering through your city streets! 🚨

**CO2 Emissions per MWh** – Let’s see how much carbon’s being emitted for every megawatt-hour of power. (The more we know, the greener we grow! 🌳)

**Now, let’s clean up some fuel names and make sure everything aligns:**

```{r}
NTD_EMISSIONS <- NTD_EMISSIONS |> 
  rename(
    "Bio diesel" = "Bio-Diesel",
    "Natural Gas" = "C Natural Gas",
    "LNG" = "Liquified Nat Gas",
    "LPG" = "Liquified Petroleum Gas",
    "Battery Electric" = "Electric Battery",
    "Electric" = "Electric Propulsion"
  )
```

### The Formula – It’s Math, But Make It Eco! 🧑‍🏫

Now for the fun part – the math! 💡

We’re going to use a long-ish arithmetic formula (don’t worry, it’s just a couple of steps) inside a mutate function to compute the total emissions for each Agency + Mode Pair. We’ll need to convert combustion fuels to CO2 emissions – but don’t panic, we’ve got conversion factors! 💪 You can hard code these bad boys in your script for simplicity:

```{r}
fuel_emission_factors <- c(
  "Diesel Fuel" = 10.19,    # kg CO2 per gallon
  "Gasoline" = 8.89,
  "Natural Gas" = 52.91,    # kg CO2 per MMBtu
  "LPG" = 5.79,  
  "LNG" = 4.29,
  "Kerosene" = 9.96,
  "Ethanol" = 1.94,
  "Hydrogen" = 0,           # No emissions from hydrogen
  "Battery Electric" = 0,   # Handled separately using CO2_MWh
  "Electric" = 0            # Handled separately using CO2_MWh
)
```

**Then, apply the emissions calculation:**

```{r}
NTD_EMISSIONS <- NTD_EMISSIONS |>
  mutate(
    Fuel_Emissions = (`Diesel Fuel` * fuel_emission_factors["Diesel Fuel"]) +
      (`Gasoline` * fuel_emission_factors["Gasoline"]) +
      (`Natural Gas` * fuel_emission_factors["Natural Gas"]) +
      (`LPG` * fuel_emission_factors["LPG"]) +
      (`LNG` * fuel_emission_factors["LNG"]) +
      (`Kerosene` * fuel_emission_factors["Kerosene"]) +
      (`Ethanol` * fuel_emission_factors["Ethanol"]) +
      (`Hydrogen` * fuel_emission_factors["Hydrogen"])
  )

NTD_EMISSIONS <- NTD_EMISSIONS |>
  mutate(
    Electricity_Emissions = `Electric` * CO2_MWh
  )

NTD_EMISSIONS <- NTD_EMISSIONS |>
  mutate(
    Total_Emissions = Fuel_Emissions + Electricity_Emissions
  )
```

**💚 Let’s Save the Planet – One Data Row at a Time! Now that our emissions data is ready, let's take a look at the final product! 🌍**

```{r}
NTD_EMISSIONS |> select(Agency, Mode, State, Fuel_Emissions, Electricity_Emissions, Total_Emissions) |> arrange(desc(Total_Emissions))
```

## Normalize Emissions to Transit Usage 🚆

In this task, we’re on a mission to normalize emissions (a fancy way of saying "make fair comparisons") based on:

**Emissions per Unlinked Passenger Trip (UPT) 🚌👥**

**Emissions per Passenger Mile 🚋🌱**

### Normalize Emissions – Leveling the Playing Field! 🎯

Before we start handing out gold medals 🏅 (or shameful side-eyes 👀), we normalize emissions based on ridership. This ensures a fair fight—because comparing a tiny rural bus route to New York City’s subway isn’t exactly fair, right? 😆

```{r}
# Filter out agencies with extremely small values to avoid distortions

AGENCY_NORMALIZED <- NTD_SERVICE %>%   # Use your actual dataset
  left_join(NTD_EMISSIONS, by = "NTD ID") %>%
  filter(!is.na(Total_Emissions))  # Ensure you have valid emission data

# Filter out agencies with extremely small values to avoid distortions
AGENCY_NORMALIZED <- AGENCY_NORMALIZED %>%
  filter(UPT > 1000, MILES > 1000) %>%  # Adjust thresholds as needed
  mutate(
    Emissions_per_UPT = ifelse(UPT > 0, Total_Emissions / UPT, NA),
    Emissions_per_Passenger_Mile = ifelse(MILES > 0, Total_Emissions / MILES, NA)
  )
```

🚦 Now, with that done, let’s move on to **finding our champions!**

## **🥇 The Greenest Transit Agency (GTA IV Award #1) 🌱🚆**

Which agency has the **lowest emissions per passenger trip**? And which one produces the least pollution **per mile traveled**?

Let's **roll out the green carpet** and announce the winners! 🎤🥁

```{r}
AGENCY_NORMALIZED <- AGENCY_NORMALIZED %>%
  mutate(Size_Category = case_when(
    UPT < 10^6 ~ "Small",
    UPT >= 10^6 & UPT < 10^8 ~ "Medium",
    UPT >= 10^8 ~ "Large"
  ))
```

### Find the greenest transit agency by lowest emissions per UPT
```{r}
AGENCY_NORMALIZED <- AGENCY_NORMALIZED %>%
  rename(Agency = Agency.x)  # Rename "Agency.x" to "Agency"

greenest_agency_UPT <- AGENCY_NORMALIZED %>%
  filter(!is.na(Size_Category) & Emissions_per_UPT > 0) %>%
  group_by(Size_Category) %>%
  slice_min(Emissions_per_UPT, n = 1, with_ties = FALSE) %>%
  select(Size_Category, Agency, Emissions_per_UPT)
kable(greenest_agency_UPT, caption = "Greenest Agency By lowest Emissions Per UPT")
```

### Find the greenest transit agency by lowest emissions per Passenger Mile

```{r}
greenest_agency_mile <- AGENCY_NORMALIZED %>%
    filter(!is.na(Size_Category) & Emissions_per_UPT > 0) %>%
  group_by(Size_Category) %>%
  slice_min(Emissions_per_Passenger_Mile, n = 1, with_ties = FALSE) %>%
  select(Size_Category, Agency, Emissions_per_Passenger_Mile)
kable(greenest_agency_mile, caption= "Greenest Transit Agency By Lowest Emissions Per Passenger Mile")
```

## **Total Emissions Avoided (GTA IV Award #2) 🌱🚆**

Now, let’s compute the **total emissions avoided** by public transit. Because without it, we'd have **millions of solo drivers** making traffic even worse than it already is! 😱🚗🚕🚙

How do we do this?\
1️⃣ **Estimate emissions if all transit passengers drove cars instead.**\
2️⃣ **Use US CAFE fuel standards to convert miles to gasoline usage.**\
3️⃣ **Calculate emissions avoided by taking public transit instead of driving.**

**Used average CAFE standard of 24.7 mpg and CO2 emissions of 8.89 kg per gallon**

```{r}
AGENCY_NORMALIZED <- AGENCY_NORMALIZED %>%
  mutate(
    Gallons_of_fuel = MILES / 24.7,  # Convert miles to fuel gallons
    Emissions_from_driving = Gallons_of_fuel * 8.89,  # Convert gallons to CO2
    Emissions_avoided = Emissions_from_driving - Total_Emissions  # Calculate CO2 saved
  )
```

### Find agency with most emissions avoided 🏆

```{r}
most_emissions_avoided_by_size <- AGENCY_NORMALIZED %>%
  filter(!is.na(Size_Category)) %>%  # Ensure valid categories
  group_by(Size_Category) %>%
  slice_max(Emissions_avoided, n = 1, with_ties = FALSE) %>%
  select(Size_Category, Agency, Emissions_avoided)
kable(most_emissions_avoided_by_size, caption = "Most Emissions Avoided")

```

This agency has officially **saved the planet** more than your coworker with the reusable straw. 🌱💚♻️

### 🏆Most Improved Green Agency **(GTA IV Award #3) 🚆** 🏆

This award recognizes the transit agency that has demonstrated the most significant improvement in environmental efficiency compared to its peers. Instead of using past-year data (which may not always be available), this award is given to the agency that has the largest improvement relative to the median emissions per passenger mile in its category (Large, Medium, or Small).

```{r}

most_improved_agency <- AGENCY_NORMALIZED %>%
  filter(!is.na(Emissions_per_Passenger_Mile) & Emissions_per_Passenger_Mile > 0) %>%
  group_by(Size_Category) %>%
  mutate(median_emissions = median(Emissions_per_Passenger_Mile, na.rm = TRUE)) %>%
  mutate(emissions_improvement = median_emissions - Emissions_per_Passenger_Mile) %>%
  slice_max(order_by = emissions_improvement, n = 1) %>%  # Select top in each category
  ungroup() %>%
  select(Agency, Size_Category, Emissions_per_Passenger_Mile, median_emissions, emissions_improvement)

kable(most_improved_agency, caption = "Most Improved Agency")
```

### 💀Biggest Gas Guzzler⛽ **(GTA IV Award #4) 🌱🚆**

This award highlights the least fuel-efficient transit agencies in each size category (Large, Medium, Small). The "Biggest Gas Guzzler" is the agency with the highest Emissions per Passenger Mile, indicating it produces the most pollution per mile traveled by a passenger. This award serves as a benchmark for identifying areas where fuel efficiency improvements and sustainability measures are most needed.

```{r}
biggest_gas_guzzler_by_size <- AGENCY_NORMALIZED %>%
  filter(!is.na(Size_Category)) %>%
  group_by(Size_Category) %>%
  slice_max(Emissions_per_Passenger_Mile, n = 1, with_ties = FALSE) %>%
  select(Size_Category, Agency, Emissions_per_Passenger_Mile)

kable(biggest_gas_guzzler_by_size, caption ="Biggest Gas Guzzler")
```

## 📈 Visualizations 🎨

#### 🚍 Greenest Transit Agency - A Breath of Fresh Air! 🌍

```{r}
greenest_data <- data.frame(
  Agency = c("New Jersey Transit Corporation", "King County, dba: King County Metro", "Altoona Metro Transit, dba: AMTRAN"),
  Size_Category = c("Large", "Medium", "Small"),
  Emissions_per_UPT = c(0.0024014936, 0.0007222026, 0.0056772412),
  Emissions_per_Passenger_Mile = c(0.0002060647, 0.0001889437, 0.0015635784)
) %>% pivot_longer(cols = c(Emissions_per_UPT, Emissions_per_Passenger_Mile), 
                    names_to = "Metric", values_to = "Emissions")

# Adjusting graph to make sure bars are visible
ggplot(greenest_data, aes(x = reorder(Agency, Emissions), y = Emissions, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +  # Increased bar width
  geom_text(aes(label = Size_Category), position = position_dodge(width = 0.7), vjust = -0.5, color = "black", size = 4) +  # Increased text size
  labs(title = "Greenest Transit Agency", y = "Emissions (kg CO₂)", fill = "Metric") +
  scale_y_continuous(expand = c(0, 0), limits = c(0, max(greenest_data$Emissions) + 0.002)) +  # Adjust y-axis to ensure bars are visible
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Talk about making a clean getaway! The MTA Bus Company and its smaller counterparts are taking the "green" route. We're not just talking about eco-friendly buses, but real reductions in emissions. A round of applause for the greenest of them all!

#### 🌱 Most Emissions Avoided - The Green Warriors 💨
```{r}
# Define the emissions data
emissions_data <- data.frame(
  Agency = c("MTA New York City Transit", "Metro-North Commuter Railroad Company", "Hampton Jitney, Inc."),
  Emissions_Avoided = c(3441010874, 413801729, 13333018)
)
# Create the plot
ggplot(emissions_data, aes(x = reorder(Agency, Emissions_Avoided), y = Emissions_Avoided, fill = Agency)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Most Emissions Avoided", y = "CO₂ Avoided (kg)", x = "Agency") +
  theme_minimal() +
  scale_y_log10(labels = scales::comma) +  # Apply logarithmic scale to the y-axis
  geom_text(aes(label = scales::comma(Emissions_Avoided)), hjust = -0.1, size = 2)  # Adjust text positioning and size


```

These agencies are saving the planet one ride at a time! Hudson Transit Lines and its pals are on a mission to leave the world with fewer emissions. They're all about that "low-carbon footprint" lifestyle!

#### 🚀 Most Improved Green Agency - Turning Green Around! 🌿

```{r}
## Most Improved Green Agency (Dumbbell Chart)
library(ggalt)
improved_data <- data.frame(
  Agency = c("MTA Bus Company", "Champaign-Urbana Mass Transit", "City of Decatur Transit"),
  Before = c(0.1260740,0.2200267, 0.3648839),
  After = c(0.0002060647,0.0001889437, 0.0015635784)
)

ggplot(improved_data, aes(y = Agency, x = Before, xend = After)) +
  geom_dumbbell(color = "blue", size = 4, dot_guide = TRUE, dot_guide_size = 0.5) +
  labs(title = "Most Improved Green Agency", x = "Emissions per Mile (kg CO₂)", y = "Agency") +
  theme_minimal()
```

These agencies took the emissions "downshift" seriously! The MTA Bus Company may have started off with high emissions, but it’s now a model of green efficiency. Talk about a green makeover!

#### 🛢️ Biggest Gas Guzzler - Oops, It’s Guzzling More than Gas! 🍃

```{r}
guzzler_data <- data.frame(
  Agency = c("Chicago Transit Authority", "WVU Personal Rapid Transit", "Pennsylvania DOT"),
  Emissions = c(162.6084, 645.1, 273.9)
)

ggplot(guzzler_data, aes(x = reorder(Agency, -Emissions), y = Emissions, fill = Emissions)) +
  geom_bar(stat = "identity") +
  scale_fill_gradient(low = "yellow", high = "red") +
  labs(title = "Biggest Gas Guzzler", y = "Emissions per Mile (kg CO₂)", x = "Agency") +
  theme_minimal()
```

Some agencies need to take a carbon diet! Looks like WVU Personal Rapid Transit took the term "gas guzzler" to the extreme. Maybe they should swap out those heavy engines for some electric vibes.

------------------------------------------------------------------------

## 📢 Conclusion

🚆 Public transit plays a *huge* role in reducing emissions! Our analysis highlights the best (and worst) performers. Keep supporting clean transit choices! 🌍💚

------------------------------------------------------------------------

## 📚 References

-   National Transit Database (NTD)
-   Energy Information Administration (EIA)
-   Environmental Protection Agency (EPA)
